---
title: "FishyG — Chinook survey sites with daily weather layers"
format: html
editor: visual
---

Extends `Linnea_folder/shinyapp.qmd`: the same survey-site map and model filter,
plus daily temperature and precipitation read straight out of
`data/Global Weather/merged_tavg_prcp_1960_2025.nc`.

The netCDF holds two griddable variables, `tavg` and `pr`. A `SpatRaster` is one
variable, so `rast()` is called twice with `subds =` rather than once. See
`HANDOFF.md` for how the file was built.

## Setup

```{r load-packages, echo = FALSE, message = FALSE, warning = FALSE}
library(shiny)
library(leaflet)
library(terra)
library(readr)
library(dplyr)

# Only the packages this app actually uses. Linnea's version loads ~25; the rest
# were unused and each one slows app start.

if (utils::packageVersion("leaflet") < "2.2.0") {
  warning("leaflet < 2.2.0 cannot draw a SpatRaster directly. ",
          "Upgrade, or wrap rasters in raster::raster() before addRasterImage().")
}
```

```{r paths, echo = FALSE, message = FALSE}
# Walk up from the working directory until the repo's data/ folder appears, so
# this runs the same whether it's knitted from Zoe_folder or the repo root.
repo_root <- local({
  d <- normalizePath(getwd(), winslash = "/")
  root <- NA_character_
  for (i in 1:6) {
    if (dir.exists(file.path(d, "data", "Global Weather"))) { root <- d; break }
    d <- dirname(d)
  }
  if (is.na(root)) stop("Could not locate the repo's data/ directory from ", getwd())
  root
})

# NB: the folder is "Global Weather" - a space and a capital W.
nc_path  <- file.path(repo_root, "data", "Global Weather",
                      "merged_tavg_prcp_1960_2025.nc")
csv_path <- file.path(repo_root, "data", "survey_sites_vancouver_isl.csv")
```

```{r sites, echo = FALSE, message = FALSE}
MODELLED <- c("NANAIMO RIVER", "COWICHAN RIVER",
              "LITTLE QUALICUM RIVER", "CHEMAINUS RIVER")

river_coords <- read_csv(csv_path, show_col_types = FALSE) |>
  mutate(
    model = if_else(stream_name_pse %in% MODELLED, "Model", "No model"),
    # -989898 is the source's missing-data sentinel, not a count.
    stream_observed_count = na_if(stream_observed_count, -989898)
  )

site_years <- sort(unique(river_coords$year))
```

```{r rasters, echo = FALSE, message = FALSE}
# One SpatRaster per variable. Neither call loads pixels: terra keeps the file
# on disk and reads a layer only when one is asked for, which is what makes a
# 24,104-layer daily slider workable.
tavg_r <- rast(nc_path, subds = "tavg")
pr_r   <- rast(nc_path, subds = "pr")

# An unset CRS is the usual reason a raster lands in the wrong hemisphere with
# no error. The file is CF-1.8 with degrees_north/east, so terra normally infers
# this, but assert it rather than hope.
if (!nzchar(crs(tavg_r))) crs(tavg_r) <- "EPSG:4326"
if (!nzchar(crs(pr_r)))   crs(pr_r)   <- "EPSG:4326"

# The dates the app is allowed to show come from the file, never from seq().
# Daymet uses a 365-day calendar, so 31 December is absent in 2016, 2020 and
# 2024: there are 24,104 layers where a naive date sequence gives 24,107.
nc_dates <- as.Date(terra::time(tavg_r), tz = "UTC")
stopifnot(!anyNA(nc_dates), length(nc_dates) == nlyr(tavg_r))

nc_years <- sort(unique(as.integer(format(nc_dates, "%Y"))))

# Resolve a date to a layer by MATCHING, never by arithmetic. "Days since
# 1960-01-01, plus one" drifts one layer late after 2016, two after 2020 and
# three after 2024 - showing plausible weather from the wrong day, silently.
layer_for <- function(d) {
  i <- match(as.Date(d), nc_dates)
  if (is.na(i)) NULL else i
}
```

```{r palettes, echo = FALSE, message = FALSE}
# Domains are fixed across the whole record, so days stay comparable as the
# slider moves. Measured once over all 24,104 layers rather than recomputed at
# every app start, which would read ~130 MB.
TAVG_DOMAIN <- c(-22.7, 32.8)   # degC
PR_MAX      <- 249.1            # mm/day

# Daily precipitation is zero-inflated (46% of cell-days are under 0.1 mm) and
# heavily right-skewed (99th percentile 49.5 mm, max 249 mm). A linear ramp to
# the wettest day on record would render almost every day as the same colour,
# so bin it on roughly logarithmic breaks instead.
PR_BREAKS <- c(0, 0.1, 1, 2.5, 5, 10, 20, 35, 50, PR_MAX)

# na.color transparent: five southwest cells lie outside the PNWNAmet domain and
# are NaN throughout. The default would paint grey squares over open water.
pal_tavg <- colorNumeric("RdYlBu", domain = TAVG_DOMAIN,
                         na.color = "transparent", reverse = TRUE)
pal_pr   <- colorBin("YlGnBu", domain = c(0, PR_MAX), bins = PR_BREAKS,
                     na.color = "transparent")

pal_model <- colorFactor(c("#1b9e77", "#7570b3"),
                         domain = c("Model", "No model"))

VAR_GROUPS  <- c("Temperature", "Precipitation")
SITE_GROUP  <- "Survey sites"

# Fit to the weather grid, not to the sites: the survey data reaches Haida Gwaii
# and up the mainland coast, while the raster covers eastern Vancouver Island
# only. Sites outside the grid are still drawn - there is just no weather under
# them.
grid_ext <- ext(tavg_r)
```

## App

```{r app, echo = FALSE}
ui <- fluidPage(
  leafletOutput("map", height = 640),
  absolutePanel(
    bottom = 60, left = 30, width = 290,
    draggable = TRUE,
    style = "background: rgba(255,255,255,0.94); padding: 14px;
             border-radius: 5px; box-shadow: 0 0 8px rgba(0,0,0,0.25);",

    tags$h4("Site map", style = "margin-top:0"),

    checkboxGroupInput("model", "Model:",
                       choices  = c("Model", "No model"),
                       selected = c("Model", "No model")),

    radioButtons("layer", "Weather layer:",
                 choices  = c("Temperature", "Precipitation"),
                 selected = "Temperature"),

    selectInput("year", "Year:",
                choices  = nc_years,
                selected = 2020),

    # A single slider over all 24,104 days would give about one pixel per 100
    # days. Scoping it to the selected year keeps it at <= 365 positions, where
    # dragging is meaningful and the arrow keys move exactly one day.
    sliderInput("day", "Date:",
                min   = as.Date("2020-01-01"),
                max   = as.Date("2020-12-31"),
                value = as.Date("2020-07-01"),
                step  = 1, timeFormat = "%d %b %Y",
                animate = animationOptions(interval = 400, loop = FALSE)),

    checkboxInput("sites_this_year", "Only sites surveyed this year", FALSE),

    tags$p("Weather grid covers eastern Vancouver Island only.",
           style = "font-size:11px; color:#666; margin-bottom:0")
  )
)

server <- function(input, output, session) {

  # Rebuild the day slider from the dates actually present in the chosen year.
  # This is also what stops 31 Dec 2016/2020/2024 from being selectable: those
  # days are not in the record, so they never become the slider's max.
  observeEvent(input$year, {
    yd <- nc_dates[format(nc_dates, "%Y") == input$year]
    keep <- if (!is.null(input$day) &&
                format(input$day, "%Y") == input$year) input$day else yd[1]
    updateSliderInput(session, "day",
                      min = min(yd), max = max(yd), value = keep)
  }, ignoreInit = FALSE)

  # Dragging across a year fires several hundred invalidations where an annual
  # slider fired one. Redraw when the drag settles instead of at every day.
  sel_date <- debounce(reactive(req(input$day)), 250)

  sites <- reactive({
    d <- river_coords[river_coords$model %in% input$model, ]
    if (isTRUE(input$sites_this_year)) {
      d <- d[d$year == as.integer(input$year), ]
    } else {
      # One marker per stream, not one per stream-year.
      d <- d[!duplicated(d$stream_name_pse), ]
    }
    d
  })

  output$map <- renderLeaflet({
    leaflet() |>
      addProviderTiles(providers$OpenStreetMap) |>
      fitBounds(xmin(grid_ext), ymin(grid_ext),
                xmax(grid_ext), ymax(grid_ext)) |>
      # Sites only. The weather layer is chosen by the radio button instead of
      # by leaflet baseGroups: a base group added to the control after the fact
      # is drawn regardless of which radio is lit, so both rasters would stack
      # until the user toggled - and it would read two layers per frame rather
      # than one.
      addLayersControl(
        overlayGroups = SITE_GROUP,
        options = layersControlOptions(collapsed = FALSE)
      )
  })

  # --- weather layer + its legend -------------------------------------------
  observe({
    req(input$layer)
    d <- sel_date()
    i <- layer_for(d)
    # A date with no layer should redraw nothing, not raise. Can only happen if
    # the slider is driven outside the range set for the year.
    if (is.null(i)) return()

    p <- leafletProxy("map") |> clearGroup(VAR_GROUPS)

    # Nearest-neighbour: the cells are ~6 km, and bilinear smoothing would imply
    # more resolution than the data has.
    # The legend carries a layerId so re-adding replaces it in place. Do NOT use
    # clearControls() here - it would also delete the layers control added in
    # renderLeaflet.
    if (input$layer == "Temperature") {
      p |>
        addRasterImage(tavg_r[[i]], colors = pal_tavg, opacity = 0.72,
                       project = TRUE, method = "ngb", group = "Temperature") |>
        addLegend("bottomright", pal = pal_tavg, values = TAVG_DOMAIN,
                  title = "Mean air temp<br/>(&deg;C)", opacity = 0.9,
                  layerId = "wxlegend")
    } else {
      p |>
        addRasterImage(pr_r[[i]], colors = pal_pr, opacity = 0.72,
                       project = TRUE, method = "ngb", group = "Precipitation") |>
        addLegend("bottomright", pal = pal_pr, values = PR_BREAKS,
                  title = "Precipitation<br/>(mm/day)", opacity = 0.9,
                  layerId = "wxlegend")
    }
  })

  # --- survey sites, drawn after the raster so they sit on top ---------------
  observe({
    d <- sites()
    p <- leafletProxy("map") |> clearGroup(SITE_GROUP)
    if (nrow(d) == 0) return()

    p |> addCircleMarkers(
      data = d, lng = ~longitude, lat = ~latitude,
      group = SITE_GROUP,
      fillColor = ~pal_model(model), fillOpacity = 0.95,
      stroke = TRUE, color = "#ffffff", weight = 1, radius = 6,
      popup = ~paste0("<b>", stream_name_pse, "</b><br/>",
                      "Model: ", model, "<br/>",
                      "Year: ", year, "<br/>",
                      "Chinook counted: ",
                      ifelse(is.na(stream_observed_count), "not recorded",
                             format(stream_observed_count, big.mark = ",")))
    )
  })
}

shinyApp(ui, server)
```

## Notes

- **Daily is weather, not climate.** This answers "what were conditions during
  this window", not "is this river warming". An annual view is `tapp()` on the
  same two rasters — worth adding as a second mode rather than replacing this.
- **The 2013 splice.** PNWNAmet before 2013, Daymet after, bias-corrected to the
  PNWNAmet baseline. The correction is a monthly-mean adjustment, so it removes
  systematic bias but not differences in day-to-day variability — which is
  exactly what a daily view puts on screen. Don't read a change in day-to-day
  texture across 2013 as a real signal.
- **Survey years are 1929–2024; the weather record is 1960–2025.** The year
  selector is driven by the raster, so pre-1960 survey years aren't reachable
  here.
- Reads are cheap (~4 ms for both variables) but the *first* one costs ~0.17 s
  while the netCDF chunk cache fills. If that stall matters, rechunk once with
  `nccopy -c time/30,lat/19,lon/31 -d 4 -s` — it takes ~2 s and the file gets
  smaller.
