---
title: "fishyG_RFModel_20260825"
format: html
editor: visual
---

## **fishyG script for building
predictive model**

For now, let's try a random forest regression model.

Here, I imagine input into the random forest model as a large dataframe\
\$1 year\
\$2 that year's precipitation until Jun30\
\$3 that year's temp until Jun30\
\$4:n average daily stream discharge\
\$n:n average daily stream temperature\
And the predicted output would be the date of first salmon returned (numeric 0:365)

#### First, load in libraries

```{r}
library(tidyverse)
library(ggplot2)
library(randomForest)
```

Maybe define some pseudo code here to later replace

```{r}
#Read in csv of data
path <- 
df <- read.csv(path) 

#reformat data into data.frame?
df %>% as.data.frame() #?
#need to rename any cols or do anything else?

#Set parameters for function
ntree <- x
mtry <- x

function{
  randomForest(salmonday ~ ., data = df, ntree = ntree, mtry = mtry)
  return() #what does this return? #can return number & MSE?
  }

```
