---
title: "fishy_G_RFModel_20260827"
format: html
editor: visual
---

## \*\*fishyG script for building a predictive model\*\*

For now, let's try a random forest regression model.

##### First, load in libraries

```{r}
library(tidyverse)
library(ggplot2)
library(randomForest)
```

##### Read in the previously generated hazard files for each stream

```{r}
path <- "/home/jovyan/Desktop/fishyG/data/stream_hazard_data/"
read.csv(paste0(path, "chemainus_hazard_data.csv")) -> chemainus
read.csv(paste0(path, "cowichan_hazard_data.csv")) -> cowichan
read.csv(paste0(path, "lilqualicum_hazard_data.csv")) -> lilqualicum
read.csv(paste0(path, "nanaimo_filteredyears_hazard_data.csv")) -> nanaimo
```

In thinking about the data & what exactly we are predicting,

#### Do RandomForest

##### Split the data 1st

```{r}
#for now, set df as cowichan to test, but this should be more reproducible
df <- cowichan
set.seed(69)
train <- sample(nrow(df), 0.7 * nrow(df))
test <- setdiff(seq_len(nrow(df)), train) 
#train & test are just row numbers of the data to use
```

##### Train the model

Predict \$arrived from flow + flow_7d + flow_14d + flow_30d + flow_delta_14d + flow_cum_since_start + days_since_high_flow + jan_jun_mean_flow

Rm flow_cum_since_start since is directly correlated with doy (data leakage), also maybe we have too many vars since n years of data is not super high... focus on vars that might directly affect whether salmon can return/swim at that moment : flow + flow_7d + days_since_high_flow + jan_jun_mean_flow

When tried to run arrived \~ all vars, got warning "The response has five or fewer unique values. Are you sure you want to do regression?" so maybe turn arrived into a factor & rerun to do classification?

```{r}
randomForest(arrived ~ flow + flow_7d + days_since_high_flow +  jan_jun_mean_flow, data = df[train,], ntree = 500)
```
